File-based localization of numerical perturbations in data analysis pipelines

### 2020-07-29 paper.tex
Update revisions

%%%% Commands %%%%
% \newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
% \newcommand{\note}[2]{\color{blue}Note from \textsc{#1}: #2\color{black}}
\newcommand{\revision}[1]{\color{purple}\textbf{REVISION:}#1\color{black}}
\newcommand{\revised}[1]{\color{blue}\textbf{REVISED:}#1\color{black}}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}
\newcommand{\toolname}[0]{Spot\xspace}
\newcommand{\flirt}[0]{\texttt{flirt}\xspace}
\newcommand{\fnirt}[0]{\texttt{fnirt}\xspace}

@@

\revised{
The labeled graph can be different depending on the condition that we obtained \reprozip trace as it can produce 
different DAGs. Even using the same DAGs, it depends on the order of executions in which condition 
we execute pipeline to pinpoint the propagation of differences. 
Therefore, any permutation of these two factors can produce different results.
% Pinpointing differences in one condition can be different than other condition.

% add a paragraph to the methods, to explain that the order matters, due to:
% 1. possible differences in the reprozip trace
% 2. the way the comparison is done
}

\begin{figure*}
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{figures/p-graph.pdf}
    \caption{Raw provenance graph (\reprozip output), with transient files shown in gray boxes.}
    \label{fig:provenance-graph}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.7\linewidth]{figures/p-graph-dag.pdf}
    \caption{Provenance DAG, with disambiguated transient files.}
    \label{fig:provenance-dag}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
      \centering
      \includegraphics[width=.7\linewidth]{figures/p-graph-dag-labelled.pdf}
     \caption{Labeled DAG comparing 2 execution conditions, showing 1 non-reproducible process.}
     \label{fig:labeled-dag} 
  \end{subfigure}
    % \includegraphics[width=.8\columnwidth]{figures/provenance-graphs}
    \caption{Provenance graphs created from the example pipeline in
    Listing~\ref{listing:sample-script}. Processes are represented with
    circles, files with rectangles, and read/write accesses with plain edges. For convenience, the process tree 
    is also shown, with gray dashed edges. Processes forked by
    \texttt{bet} were captured by \reprozip while they did not appear in
    Listing~\ref{listing:sample-script}. Processed associated with
    executables located in \texttt{/usr/bin/} or \texttt{/bin/} are not shown.
    \revised{Node labels are enlarged.}}
    \label{fig:spot-example}
  \end{figure*}

@@

We use \reprozip~\cite{rampin2016reprozip}
to capture: (1) the set of processes created by the
pipeline, and
(2) the set of files read and written by each process, including
temporary files. \reprozip collects this information through the
\texttt{ptrace()} system call, with no required instrumentation of the pipeline.
Using the \reprozip trace, \toolname reconstructs a provenance graph by creating process and file
nodes and by adding directed edges corresponding
to file reads and writes (Figure~\ref{fig:provenance-graph}).

@@

@@

We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2
CentOS versions. 
\revised{Using the \reprozip trace obtained from CentOS7, 
we captured transient files in CentOS6 and labeled pipeline processes in CentOS7.}
Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We compared pipeline results using  FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff}, to ignore execution-specific information such as file path or
timestamps. To compare segmentations $X$ and $Y$, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\]

\revised{
The Dice coefficient~\cite{dice1945measures} is a commonly used metric in validating medical image segmentation.
The Dice value ranges from 0 to 1, with 1 indicating a complete overlap between two segmentation results.
Besides, the Jaccard coefficient~\cite{jaccard1912distribution} can be considered an equivalent metric 
to measure the overlap between images.

% Also, we used the Pearson correlation coefficient to establish the relationship between Dice value and segmented region size.
% Like other correlation coefficients, this one varies between -1 and +1 with 0 implying no correlation. 
% Correlations of -1 or +1 imply an exact linear relationship. 
}

@@

\section{Results}
\revised{
All experiments were run on the machine with a 3.4GHz Intel Core i7 processor 
and 32GB memory under the CentOS7 operating system.
The average processing time per subject was approximately 2~hours for
PreFreeSurfer and 8~hours for FreeSurfer. The average output file size was
2.7~GB for PreFreeSurfer and 4.1~GB for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}
For each subject, PreFreeSurfer accessed
83K files and created 7.7K processes, and FreeSurfer accessed 62K files and 
created 4K processes.
The scripts and analyses used to create the figures in this section are available at 
\url{https://github.com/big-data-lab-team/HCP-reproducibility-paper}.
}
% \subsection{Subject clustering}

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps.
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  fraction of subjects for which the corresponding process wasn't reproducible. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. 
  \revised{Grey cells show that any command line of the process did not occur in the pipeline step.}}
  \label{fig:pfs_heatmap}
\end{figure*}

@@

\revised{
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject using the checkerboard pattern.
Differences appear to be visually important, in particular in the areas framed in red.
The checkerboard representation is a common method to illustrate the magnitude of the differences in registration results.
Although differences are more clearly represented in the animations linked in this figure.}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that Dice coefficents below 0.9 are 
observed in most regions, and particularly in the smallest ones.
\revised{
The correlation between the Dice values and region sizes computed as the Pearson's correlation coefficient 
was 0.12 (\emph{P}=.43). 
The positive correlation implies that as region size increases, Dice value increases. 
% This is an average Dice values over 20 subjects for each 44 regions.
}

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png} 
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template. 
  \revised{Each voxel shows the number of subjects that created different results. 
  The color bar ranges from 1 to 7 as the minimum and the maximum number of subjects in each voxel.
  Voxels are not colored if the identical results were produced in all 20 subjects.}
  An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here}.} 
  \label{fig:tissue_class}
\end{figure}

@@

To evaluate the reproducibility of a pipeline, \toolname needs to execute
it 3 times in order to (1) record a first \reprozip trace, (2) save
transient files, and (3) compare results in the second condition. This is
one more execution than the theoretical minimum of 2. It might be possible
to further reduce this overhead by executing at step (2) only the
processes depending on transient files.

>>>>

%%%% Commands %%%%
% \newcommand{\todo}[1]{\color{red}\textbf{TODO:}#1\color{black}}
% \newcommand{\note}[2]{\color{blue}Note from \textsc{#1}: #2\color{black}}
\newcommand{\revision}[1]{\color{purple}\textbf{REVISION:}#1\color{black}}
\newcommand{\revised}[1]{\color{blue}#1\color{black}}
\newcommand{\reprozip}[0]{ReproZip\xspace}
\newcommand{\tristan}[1]{\color{orange}\textbf{From Tristan:}#1\color{black}}
\newcommand{\toolname}[0]{Spot\xspace}
\newcommand{\flirt}[0]{\texttt{flirt}\xspace}
\newcommand{\fnirt}[0]{\texttt{fnirt}\xspace}

@@

\begin{figure*}
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.9\linewidth]{figures/p-graph.pdf}
    \caption{Raw provenance graph (\reprozip output), with transient files shown in gray boxes.}
    \label{fig:provenance-graph}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
    \centering
    \includegraphics[width=.7\linewidth]{figures/p-graph-dag.pdf}
    \caption{Provenance DAG, with disambiguated transient files.}
    \label{fig:provenance-dag}
  \end{subfigure}
  \hfill
  \begin{subfigure}[t]{0.3\linewidth}
      \centering
      \includegraphics[width=.7\linewidth]{figures/p-graph-dag-labelled.pdf}
     \caption{Labeled DAG comparing 2 execution conditions, showing 1 non-reproducible process.}
     \label{fig:labeled-dag} 
  \end{subfigure}
    % \includegraphics[width=.8\columnwidth]{figures/provenance-graphs}
    \caption{Provenance graphs created from the example pipeline in
    Listing~\ref{listing:sample-script}. Processes are represented with
    circles, files with rectangles, and read/write accesses with plain edges. For convenience, the process tree 
    is also shown, with gray dashed edges. Processes forked by
    \texttt{bet} were captured by \reprozip while they did not appear in
    Listing~\ref{listing:sample-script}. Processed associated with
    executables located in \texttt{/usr/bin/} or \texttt{/bin/} are not shown.}
    \label{fig:spot-example}
  \end{figure*}

@@

We use \reprozip~\cite{rampin2016reprozip}
to capture: (1) the set of processes created by the
pipeline, and
(2) the set of files read and written by each process, including
temporary files. \reprozip collects this information through the
\texttt{ptrace()} system call, with no required instrumentation of the pipeline.
Using the \reprozip trace, \toolname reconstructs a provenance graph by creating process and file
nodes and by adding directed edges corresponding
to file reads and writes (Figure~\ref{fig:provenance-graph}).
\revised{We assume that provenance graphs are identical for the \reprozip traces obtained from 
the same subjects in different operating systems.}

@@

\revised{
The labeled graph can be different depending on the order of executions in which condition 
we execute pipeline to pinpoint the propagation of differences. 
Therefore, we label processes in two order of executions, and then aggregate findings so that label processes 
as non-reproducible (red) if it creates different results in both order of executions.
}

@@

% \revision{Using the \reprozip trace obtained from CentOS7, 
% we captured transient files in CentOS6 and labeled pipeline processes in CentOS7.}
We processed the 20 subjects with PreFreeSurfer and FreeSurfer, using the 2
CentOS versions. 
Each subject was processed twice on the same operating
system to detect within-OS variability coming from pseudo-random
operations. We compared pipeline results using  FreeSurfer tools \texttt{mri\_diff},
\texttt{mris\_diff}, and \texttt{lta\_diff}, to ignore execution-specific information such as file path or
timestamps. To compare segmentations $X$ and $Y$, we used the Dice coefficient defined as follows:
\[DICE=\frac{2|X \cap Y|}{|X| + |Y|}\]

\revised{
The Dice coefficient~\cite{dice1945measures} is a commonly used metric in validating medical image segmentation.
The Dice value ranges from 0 to 1, with 1 indicating a complete overlap between two segmentation results.
Besides, the Jaccard coefficient~\cite{jaccard1912distribution} can be considered an equivalent metric 
to measure the overlap between images.
}

@@

\section{Results}
\revised{
All experiments were run on a machine with a 3.4GHz Intel Core i7 processor and 32GB memory 
under the CentOS7 operating system. 
The details of the computation of the PreFreeSurfer and FreeSurfer pipelines are illustrated 
in~Table~\ref{table:pipeline-stats}.
The scripts and analyses used to create the figures in this section are available at 
\url{https://github.com/big-data-lab-team/HCP-reproducibility-paper}.

\begin{table*}[ht]
  \centering
  \begin{threeparttable}
  \caption{The computation details of the pipelines per subject.}
  \label{table:pipeline-stats}
  \begin{tabular}{ccc|cc}
  \toprule
  \multicolumn{1}{c}{} & \multicolumn{2}{c}{PreFreeSurfer } &   \multicolumn{2}{c}{FreeSurfer } \\

   &        \makecell{Mean}   &  \makecell{Standard error}  & \makecell{Mean}   &  \makecell{Standard error}   \\ \midrule
  Processing time (mins)      &     106.67     &     2.68     &     nd     &     nd     \\
  Output file size (GB)       &     2.8        &     0.10     &     nd     &     nd     \\
  Number of file accesses     &     94089      &     2644.99  &     nd     &     nd     \\
  Number of processes         &     8731       &     198.18   &     nd     &     nd     \\
  \bottomrule
  \end{tabular}
  \end{threeparttable}
  \end{table*}
}
% All experiments were run on a machine with a 3.4GHz Intel Core i7 processor 
% and 32GB memory under the CentOS7 operating system.
% The average processing time per subject was approximately 2~hours (with standard error X) 
% for PreFreeSurfer and 8~hours (with standard error X) for FreeSurfer. 
% The average output file size was 2.7~GB (with standard error Y) for PreFreeSurfer 
% and 4.1~GB (with standard error Y) for FreeSurfer.
%\tristan{Ali, are these numbers per subject or for 20 subjects? it's an average of 20 subjects}
% For each subject, PreFreeSurfer accessed 83K files and created 7.7K processes, and FreeSurfer accessed 62K files and 
% created 4K processes.

% \subsection{Subject clustering}

@@

Figure~\ref{fig:pfs_heatmap} shows the frequency of non-reproducible pipeline processes
in PreFreeSurfer. 
The processes identified as non-reproducible were observed in linear registration 
with FSL \flirt (in ACPC-Alignment, Brain Extraction, Distortion Correction, and
Atlas Registration), in non-linear registration with FSL \fnirt (in Brain Extraction 
and Atlas Registration), and in image warping with FSL \texttt{new\_invwarp} (in Brain Extraction 
and Atlas Registration). Differences were also observed in image mean 
computations with FSL \texttt{maths}  (in Anatomical Average). 
Figure~\ref{fig:complete_pfs} shows a complete PreFreeSurfer labeled DAG, localizing 
the observed differences in the entire pipeline, for a given subject.
\begin{figure*}
\centering
  \includegraphics[width=\textwidth]{./figures/heatmap.pdf}
  \caption{Heatmap of non-reproducible processes across PreFreeSurfer pipeline steps.
  Each cell represents the occurrence of a particular command line in a
  pipeline step among Anatomical Average (AAve), Anterior/Posterior
  Commissure Alignment (ACPC-A), Brain Extraction (BExt), Bias Field
  Correction (BFC), or Atlas-Registration (AR). Cell labels indicate the
  fraction of subjects for which the corresponding process wasn't reproducible. For example,
  the \flirt tool was invoked 6 times in step DC for each of the 20
  subjects: 2 instances weren't reproducible in 19 subjects, 3
  instances were always reproducible, and 1 instance wasn't reproducible in
  17 subjects. 
  \revised{Grey cells show that the process did not occur in the pipeline step.}}
  \label{fig:pfs_heatmap}
\end{figure*}

@@

\revised{
Figure~\ref{fig:fnirt_result} compares \fnirt results in Brain Extraction
for a particular subject using the checkerboard pattern, 
a common method to illustrate the magnitude of the differences in registration results.
Differences appear to be visually important, in particular in the areas framed in red.}

@@

However, FreeSurfer results still differ between conditions, due to the
propagation of differences created in PreFreeSurfer. We
observed the effect of this propagation in FreeSurfer results, as shown in
Figure~\ref{fig:tissue_class} for whole-brain segmentations. The
Dice coefficients associated with the 44 regions segmented by
FreeSurfer are shown in Figure~\ref{fig:scatter_plot}, showing that Dice coefficents below 0.9 are 
observed in most regions, and particularly in the smallest ones.
\revised{However, there is no significant correlation between Dice value and region size.}

\begin{figure}
%  \includegraphics{brain\_classification}
\centering
  \includegraphics[width=\columnwidth]{figures/brain_segmentation_mni.png} 
  \caption{Sum of binarized differences between whole-brain FreeSurfer
  segmentations obtained from PreFreeSurfer processings in CentOS6 vs CentOS7
   (N=20). Segmentations were resampled and overlaid to the MNI152 volume
  template. 
  \revised{
  The binary images are created based on the intensity differences between result images in CentOS6 and CentOS7. 
  The voxel is 1 if the intensity value of the corresponding voxel in the images is different, otherwise it is 0.
  Each voxel shows the number of subjects that created different results. 
  % The color bar ranges from 1 to 7 as the minimum and the maximum number of subjects in each voxel.
  Voxels are not colored if the identical results were produced in all 20 subjects.}
  An animated comparison of segmentations obtained for a particular subject is available
\href{https://github.com/big-data-lab-team/HCP-reproducibility-paper/blob/master/figures/fs_brain_segmentation.gif}
{here}.} 
  \label{fig:tissue_class}
\end{figure}

@@

\revision{To evaluate the reproducibility of a pipeline, \toolname needs to execute
it 3 times in order to (1) record a first \reprozip trace, (2) save
transient files, and (3) compare results in the second condition. This is
one more execution than the theoretical minimum of 2. It might be possible
to further reduce this overhead by executing at step (2) only the
processes depending on transient files.}
